### 2025-05-04 thesis/thesis.tex
Finish the section on neural approaches

The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
neural NL2SQL approaches:

\subsection{Benchmark Evolution}

\subsection{LLM-based NL2SQL Systems}

% Prompt engineering and fine tuning

>>>>

The evolution from early neural approaches to RAT-SQL emphasized the rapid advancements that happened in the research field of 
neural NL2SQL approaches in different dimensions:

\begin{enumerate}
    \item \textbf{Model Complexity} — Given the research progression from early sequence to sequence translation approaches 
          (\textsc{Seq2SQL}) towards sketch based and type augumented and graph based approaches (\textsc{TypeSQL, SQLNet, GNN})
          and syntax tree decoding emphasized by \textsc{SyntaxSQLNet}, neural approaches continously advanced in the complexity of 
          approaches that is required to beat state-of-the-art approaches in contemprorary benchmarks like \textsc{Spider}. RAT-SQL 
          presents one of the late and most complex advancements in the field of neural NL2SQL approaches with its adapted self 
          attention mechanism \citep{Seq2SQL, TypeSQL, GNN, SyntaxSQLNet, RATSQL}.
    \item \textbf{Transferability} — Each of the approaches introduced above represents a succession in terms of their transferability.
          The field of neural NL2SQL approaches significantly improved the ability for NLIDBs to generalize over the underlying database
          schemas. RAT-SQL showed the strongest cross-domain accuracy (that is benchmarked by the \textsc{Spider} benchmark). With
          standard benchmarks emerging it became easier to verify and quantify which approach had the highest transferability as
          \textsc{Spider} specifically had independent development and test datasets, preventing approaches from over-optimizing on
          training data \citep{Spider, RATSQL}.
    \item \textbf{Robustness} — As research systems advanced in complexity and shifted from raw input to output translation (Seq2SQL)
          their robustness steadily increased. Through more approaches like \textsc{SyntaxSQLNet} which utilized structured decoding,
          \textsc{IRNet} which relied on an intermediate representation and \textsc{RAT-SQL} the challenges around \textit{schema linking}
          outlined by \cite{RATSQL} have increasilingly led to more robust systems that can handle rephrasings, spelling mistakes and
          variations in natural language usage far beyond what traditional NL2SQL approaches could accomplish \cite{SyntaxSQLNet, IRNet, RATSQL}.
    \item \textbf{Query Complexity} — The performance on complex queries involving multiple tables, relying on complex aggregations,
          nested structures and joins dramatically improved over the course of the research that happened in this field. Whilst 
          \textsc{IRNet} reresents one of the first singificant advancements when it comes to the ability of neural approaches to handle
          complex queries, \textsc{RAT-SQL} still showed to outperform the intermediate representation approach introduced by \textsc{IRNet}
          by up to 10.5\% \citep{IRNet, RATSQL}.
    \item \textbf{Schema Understanding} — Whilst early approaches like Seq2SQL primarily applied reinforcement learning for end to end
          query generation \citep{Seq2SQL}, later approaches like \textsc{TypeSQL, GNN} and specifically \textsc{RAT-SQL} showed novel and state-of-
          the-art \textit{schema understanding / schema linking} capabilities, yielding the ability to accurately reason about user intent and
          traverse the database schema while generating queries \citep{TypeSQL, GNN, RATSQL}. 
\end{enumerate}

\subsection{Limitations of Neural Approaches}

Despite the dramatic \textit{accuracy, transferability} and \textit{robustness} improvements that could be observed with late neural
approaches \citep{IRNet, RATSQL}, neural approaches still suffered from serious shortcommings / unsolved challenges:

\begin{enumerate}
    \item \textbf{Training Data} — Utilizing neural networks these approaches required susbtantially more training data (ie. natural language
          paired with output SQL queries) than traditional systems which required serious efforts of data collection \citep{Spider}.
    \item \textbf{Correctness} — The inherent mismatch between neural networks and formal languages yielded cases where models produced
          invalid SQL code. Approaches like \textsc{SyntaxSQLNet} improved the tried to solve this circumstance by utilizing syntax trees
          during decoding but nonetheless syntactic correctness remained a challenge across future iterations of neural systems. \citep{SyntaxSQLNet}
    \item \textbf{Domain Language} — Despite increased \textit{transferability} characteristics neural approaches still suffered from a
          limited vocabulary and inter-domain understanding of terminology and relation between concepts which made highly domain
          specific natural language queries challenging.
    \item \textbf{Observability} — The black-box nature of neural networks made approaches relying on them, particularly the ones with complex
          architectures, hard do understand / explain in case when neural systems yielded undesirable output. 
\end{enumerate}

The introduction and advancement of early neural NL2SQL approaches led to significant advancements in the research and feasibility of NLIDBs. The
research shift started in this era established the foundations for further and more advanced machine learning approaches (specifically language
model oriented approaches )being researched. Neural approaches showed to significant improvements in performance when being paired with pre-trained
language models \citep{RATSQL} which led to further research on their applicability.

\subsection{Pre-trained Language Models}

% PLM approaches
% RESDSQL, Graphix-T5
% T5 fine-tuning etc.

% https://arxiv.org/abs/2109.05093

\subsection{Large Language Models}

% CodeS
% GPT-X based
% 
% Prompt engineering and fine tuning
